---
title: "R Notebook"
output: html_notebook
---

https://www.youtube.com/watch?v=I0VCGCnquTQ

```{r}
rm(list = ls())

library(twitteR)
library(tm)
library (ROAuth)
library(base64enc)
library(SnowballC)
library(syuzhet)

```

Access setup

```{r}
# need only be don once after caching

appname <- "gguSentiment"
api_key <- "oGR0V4nvOX6GFTsl2Tme1PakD"
api_secret_key <- "BXAXdkFqADsT3dDEnDlGYeDHgfp8NBqtYhpNySejE771H4ba8p"

access_token <- '901150381-B9XjpI50KC2kVnMzOInqfbfpVQjFKakzIo33ffJq'
access_token_secret <- 'uvm1Tk7gE4k0fWhjQBILGAoNeUdBAZazws82Sfn2oj1rw'

reqURL <- "https://api.twitter.com/oauth/request_token"
accessURL <- "https://api.twitter.com/oauth/access_token"
authURL <- "https://api.twitter.com/oauth/authorize"

consumerKey <- api_key
consumerSecret <- api_secret_key
accessToken <- access_token 
accessTokenSecret <- access_token_secret

# Use a local file ('.httr-oauth'), to cache OAuth access credentials between R sessions?
# Adding .httr-oauth to .gitignore  
# (do in console)
#setup_twitter_oauth(consumerKey, consumerSecret, accessToken, accessTokenSecret)

```

values to search

```{r}
user = "realDonaldTrump"

tags = c("#Nike")
limit = 100

```


get tweets from user

```{r}
tweets <- userTimeline (user, n=limit)
tweets.df <- twListToDF (tweets)
head(tweets.df)
```

Search tweets by hasthtag

```{r}

tweets = searchTwitter(tags, n=limit)
tweets.df <- twListToDF (tweets)
head(tweets.df)

```


Save to csv

```{r}
save_as_csv(tweets.df, 'tweets.csv', prepend_ids = TRUE, na = "",
  fileEncoding = "UTF-8")

```



```{r}
head(tweets.df$text)
```

Clean up tweet text to prepare for analysis 

```{r}
tweets.df2 <- gsub ("http.*", "", tweets.df$text)	# Remove http
tweets.df2 <- gsub ("https.*", "", tweets.df2)	# Remove https
tweets.df2 <- gsub ("#.*", "", tweets.df2)		# Remove #
tweets.df2 <- gsub ("@.*", "", tweets.df2)		# Remove @

head(tweets.df2)
```

Sentiment score analysis

R package “syuzhet” categorizes sentiment into 10 different emotions: 
Anger; Anticipation; Disgust; Fear; Joy;
Sadness; Surprise; Trust; Negative; Positive

```{r}
word.df <- as.vector (tweets.df2)		
emotion.df <- get_nrc_sentiment (word.df)
emotion.df2 <- cbind (tweets.df2, emotion.df)
head (emotion.df2)

```

Get most positive
```{r}
sentiment.value <- get_sentiment (word.df)
most.positive <- word.df [ sentiment.value == max (sentiment.value) ]
most.positive

```

Get the most negative
```{r}
sentiment.value <- get_sentiment (word.df)
most.negative <- word.df [ sentiment.value == min (sentiment.value) ]
most.negative


```

We can ask R to categorize tweets as positive, negative, or neutral, depending on value of sentiment.value

Declare positive, negative, and neutral tweets, based on sentiment values

```{r}
positive.tweets <- word.df [sentiment.value > 0]
head (positive.tweets)

negative.tweets <- word.df [sentiment.value < 0]
head (negative.tweets)

neutral.tweets <- word.df [sentiment.value == 0]
head (neutral.tweets)

```

Create a table of sentiment
```{r}
category_sentiment <- ifelse (sentiment.value < 0, 'Negative', 
  ifelse (sentiment.value > 0, 'Positive', 'Neutral' ))

table (category_sentiment)

```

